---
title: "App3 arima ts"
author: "Klaus KUEKOU"
date: "2025-12-09"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Road map for ARIMA


## Import Library

```{r, echo = FALSE}
library(forecast)
library(tseries)
library(ggplot2)
library(gridExtra)
```


## Load Dataset and Explore it 

```{r}
data("EuStockMarkets")
str(EuStockMarkets)
frequency(EuStockMarkets)
time(EuStockMarkets)[c(1, length(time(EuStockMarkets)))]
colnames(EuStockMarkets)
```
## Represent the choosen series


```{r}
CAC <- EuStockMarkets[, "CAC"]
autoplot(CAC) + 
  ggtitle("CAC Price Series") + 
  ylab("Price") + 
  xlab("Time") +
  theme_minimal()
```
Overall, the graph show that CAC prices tend to grow over time between 1991 and 1994 and also 1999. It's observe a decline between 1994 and 1996. These trends are associated to a variability.

## Log prices and returns

```{r}
pt <- log(CAC)
rt <- diff(pt)
```


```{r}

p1 <- autoplot(pt) + 
  ggtitle("Log-prices") + 
  ylab("Log-price") + 
  theme_minimal()

p2 <- autoplot(rt) + 
  ggtitle("Returns") + 
  ylab("Returns") + 
  theme_minimal()

grid.arrange(p1, p2, ncol=1)
```

## Check the stationnarity (ADF)

### Price series

```{r}
adf_price <- adf.test(CAC)
print(adf_price)
```

```{r}
adf_price <- adf.test(CAC, k=1)
print(adf_price)
```
Prices serie is not stationnary with strong evidence (p-value = 0.9876) at 95% level of confidence

### Returns


```{r}
adf_rt <- adf.test(rt)
print(adf_rt)
```

```{r}
adf_rt <- adf.test(rt, k=1)
print(adf_rt)
```

Returns are stationnary (p-value = 0.01) at 95% level of confidence

## ACF and PACF of returns 

```{r}
acf(na.omit(rt), main="ACF of returns")
pacf(na.omit(rt), main="PACF of returns")
```
There is only one significant pic at level one on ACP and one pic on PACF.

```{r}
r2t <- rt^2
acf(r2t, main="ACF of squared returns (r²t)", lag.max=40)
```
This graph indicates pics at lags 1, 2 and 3. It confirms the presence volatitly on the series.



## Estimation of potentials ARIMA


```{r}
ARIMA100 <- Arima(rt, order=c(1,0,0))
ARIMA001 <- Arima(rt, order=c(0,0,1))
ARIMA101 <- Arima(rt, order=c(1,0,1))
ARIMA107 <- Arima(rt, order=c(1,0,7))
ARIMA007 <- Arima(rt, order=c(0,0,7))
```


```{r}
models <- list(ARIMA100, ARIMA001, ARIMA101, ARIMA107, ARIMA007)
names_models <- c("ARIMA(1,0,0)", "ARIMA(0,0,1)", "ARIMA(1,0,1)", 
                  "ARIMA(1,0,7)", "ARIMA(0,0,7)")

comparison <- data.frame(
  Model = names_models,
  AIC = sapply(models, AIC),
  BIC = sapply(models, BIC),
  LogLik = sapply(models, logLik)
)
```



```{r}
comparison
```



```{r}
bestmodel = forecast::auto.arima(rt, d=0)
bestmodel
```

```{r}
coef_model <- coef(bestmodel)
print(coef_model)
```


The best model is AR(1). an increase of the last return (CAC), will lead an increase 
of present value.



```{r}
checkresiduals(bestmodel)

residuals <- residuals(bestmodel)
lb_test <- Box.test(residuals, lag=20, type="Ljung-Box", fitdf=length(coef(bestmodel)))
print(lb_test)
```


```{r}
lb_test_sq <- Box.test(residuals^2, lag=20, type="Ljung-Box")
print(lb_test_sq)
```
## some forecasted values 


```{r}
h <- 20
forecast_r <- forecast(bestmodel, h=h, level=95)

autoplot(forecast_r) + 
  ggtitle("Return forecasts (h=20 days, IC 95%)") +
  ylab("Returns") +
  xlab("Time") +
  theme_minimal()

print(forecast_r)
```



## Best model for ARIMA d = 1



```{r}
ARIMA110 <- Arima(pt, order=c(1,1,0))
ARIMA011 <- Arima(pt, order=c(0,1,1))
ARIMA111 <- Arima(pt, order=c(1,1,1))
ARIMA117 <- Arima(pt, order=c(1,1,7))
ARIMA017 <- Arima(pt, order=c(0,1,7))
```




```{r}
models <- list(ARIMA110, ARIMA011, ARIMA111, ARIMA117, ARIMA017)
names_models <- c("ARIMA(1,1,0)", "ARIMA(0,1,1)", "ARIMA(1,1,1)", 
                  "ARIMA(1,1,7)", "ARIMA(0,1,7)")

comparison <- data.frame(
  Model = names_models,
  AIC = sapply(models, AIC),
  BIC = sapply(models, BIC),
  LogLik = sapply(models, logLik)
)
```

```{r}
comparison
```


```{r}
bestmodel1 = forecast::auto.arima(pt, d=1)
bestmodel1
```


```{r}
coef_model1 <- coef(bestmodel1)
print(coef_model1)
```


```{r}
checkresiduals(bestmodel1)

residuals1 <- residuals(bestmodel1)
lb_test1 <- Box.test(residuals1, lag=20, type="Ljung-Box", fitdf=length(coef(bestmodel1)))
print(lb_test1)
```


```{r}
lb_test_sq1 <- Box.test(residuals1^2, lag=20, type="Ljung-Box")
print(lb_test_sq1)
```



##FORECAST BEST MODEL 1

```{r}
h <- 20
forecast_r1 <- forecast(bestmodel1, h=h, level=95)

autoplot(forecast_r1) + 
  ggtitle("Log price forecasts (h=20 days, IC 95%)") +
  ylab("Log Prices") +
  xlab("Time") +
  theme_minimal()

print(forecast_r1)
```

## Best model with CAC.

```{r}
bestmodel2 = forecast::auto.arima(CAC, d=1)
bestmodel2
```


## Comparison with Bestmodel1

```{r}
AIC(bestmodel1) < AIC(bestmodel2)
```
The conclusion here is that price logs must necessarily be used (AR(1)). However, ARIMA models are not suitable due to volatility issues. The models indicated are of the ARCH/GARCH type.
































































































